% !TEX root = ../main.tex

% <This version of CC is a minimalistic subset of Agda.>
This chapter gives a survey about defunctionalization transformation in non-dependently-typed systems (Section~\ref{sec:2.1}) and presents the definition the source language (CC) for dependently-typed defunctionalization (Section~\ref{sec:2.2}).

\section{Defunctionalization}
\label{sec:2.1}

Defunctionalization is a whole-program transformation that turns higher-order functional programs into first-order ones. This section explains type-preserving defunctionalization for simply-typed and polymorphic source programs with illustrating examples.

\subsection{Simply-typed programs}

Defunctionalization is based on the observation that a program has only finitely many function definitions. Therefore, if the whole program is available, we can enumerate every function and replace them with distinct constructors of an \textit{algebraic data type (ADT)} in the target language. I refer to these constructors as \textit{labels}. Defunctionalization translates function applications into calls to an auxiliary \textit{apply} function, which takes a label and the function argument, performs case analysis on the label, and returns the result of the corresponding function applied to the argument. In other words, letting $\bbracket{-}$ denote the transformation, $ \bbracket{e_1\ e_2} = apply\ \bbracket{e_1}\ \bbracket{e_2} $, where $apply\ \bbracket{e_1}\ \bbracket{e_2}$ simulates the behavior of $e_1\ e_2$ in the source language.

Suppose the source and the target language are both OCaml. Example \ref{ex:simple dfc} shows how to defunctionalize a simply typed higher-order program to a simply-typed first-order program.
\begin{exmp}
\label{ex:simple dfc}
Let the source program be:
\begin{lstlisting}
let rec map : (int -> int) -> int list -> int list =
    fun f ls -> match ls with
        | [] -> []
        | hd :: tl -> (f hd) :: (map f tl)
in
    map (fun x -> x + 1) (map (fun x:int -> x) [1])
\end{lstlisting}
It contains two first-class anonymous functions of the type \texttt{int -> int}. Defunctionalization involves defining a data type \texttt{lam} with two constructors, one for each function.
\begin{lstlisting}
(* Incr represents (fun x -> x + 1) *)
(* Id represents (fun x:int -> x) *)
type lam = Incr | Id
\end{lstlisting}
The next step is defining the auxiliary function \textit{apply}: \texttt{apply f arg} returns the result of
applying the source-program function which \texttt{f} represents to the argument \texttt{arg}. If \texttt{f} is \texttt{Incr}, it returns \texttt{arg + 1}, and similarly for case \texttt{Id}.
\begin{lstlisting}
let apply : lam -> int -> int = 
    fun f arg -> match f with
        | Incr -> arg + 1
        | Id -> arg 
\end{lstlisting}
Now, the defunctionalized target program can be obtained by replacing all the first-class functions with their corresponding labels and all the function applications with calls to \texttt{apply}.
\begin{lstlisting}
let rec map' : lam -> int list -> int list =
    fun f ls -> match ls with
        | [] -> []
        | hd :: tl -> (apply f hd) :: (map' f tl)
in
    map' Incr (map' Id [1])
\end{lstlisting}
Note that \texttt{map'} now takes first-order values of type \texttt{Lam} instead of functions.
\end{exmp}

\subsubsection{Type preservation}
Informally speaking, a type-preserving transformation turns a well-typed source program to a well-typed target program.
In Example~\ref{ex:simple dfc}, all first-class functions have the same type. Example~\ref{ex:illtype dfc} shows that defunctionalization produces ill-typed programs when first-class functions are not all in the same type.

\begin{exmp}
\label{ex:illtype dfc}
Let the source program be:
\begin{lstlisting}
let compose : (int -> bool) -> (int -> int) -> int -> bool =
    fun g f x -> g (f x)
in 
    compose (fun x -> x > 0) (fun x:int -> x) 1
\end{lstlisting}
Data type \texttt{lam} and the \texttt{apply} function are defined similarly as in Example~\ref{ex:simple dfc}.
\begin{lstlisting}
(* ToBool represents (fun x -> x > 0) *)
(* Id represents (fun x : int -> x) *)
type lam = ToBool | Id

let apply f arg = match f with
    | ToBool -> arg > 0
    | Id -> arg
\end{lstlisting}
Function \texttt{apply} is ill-typed -- it returns a \texttt{bool} in case \texttt{ToBool} and returns an \texttt{int} in case \texttt{Id}.
\end{exmp}

A minor adjustment can fix this problem: defining two \texttt{apply} functions, one for functions of type \texttt{int -> bool}, another one for functions of type \texttt{int -> int}.

\begin{lstlisting}
(* apply_int_bool : Lam -> int -> bool *)
let apply_int_bool f arg = match f with
	| ToBool -> arg > 0


(* apply_int_int : Lam -> int -> int *)
let apply_int_int f arg = match f with
	| Id -> arg

let compose' g f x = apply_int_bool g (apply_int_int f x) in
    compose' ToBool Id 1
\end{lstlisting}

In general, simply typed defunctionalization need to define a family of monomorphic \texttt{apply} functions (i.e.~one specialized function \texttt{apply\_a\_b} for each function type \texttt{a -> b} in the source program) to achieve type preservation.

\subsection{Polymorphic programs}

In languages with polymorphism and \textit{generalized algebraic data types (GADTs)}, defunctionalization can be done similarly as in simply-typed scenarios – first-class functions are replaced by labels, and applications are translated to calls into the \textit{apply} function. The difference is that we only need one polymorphic \textit{apply}, and function labels are defined as constructors of a GADT. GADTs allow data types to take type parameters, and constructors may return specific instantiations of the parameters. Suppose our target language now supports GADTs, Example \ref{ex:poly dfc} shows how to defunctionalize a polymorphic higher-order program.

\begin{exmp}
\label{ex:poly dfc}
Consider the polymorphic composition function:
\begin{lstlisting}
let compose : type a b c. (b -> c) -> (a -> b) -> a -> c =
    fun g f x -> g (f x) 
in
    compose (fun x -> x > 0) (fun x -> x) 1
\end{lstlisting}
\texttt{(fun x -> x)} is the polymorphic identity function. The type \texttt{lam} takes two type parameters, and a function of type \texttt{a -> b} in the source program translates into a label of type \texttt{(a, b) lam}. Note that the constructor \texttt{ToBool}, which corresponds to a non-polymorphic function, has instantiated parameters. 
\begin{lstlisting}
type ('a, 'b) lam = 
      ToBool : (int, bool) lam
    | Id : ('a, 'a) lam

let apply : type a b. (a, b) lam -> a -> b =
    fun f arg -> match f with
        | ToBool -> arg > 0
        | Id -> arg
\end{lstlisting}
Again, the target program is obtained by replacing first-class functions with their corresponding labels and function applications with calls to \texttt{apply}, and the result is well-typed.
\begin{lstlisting}
let compose' : type a b c. 
                 (b, c) lam -> (a, b) lam -> a -> c =
    fun g f x -> apply g (apply f x)
in 
    compose' ToBool Id 1
\end{lstlisting}
	
\end{exmp}

\subsubsection{Functions with free variables}
Defunctionalization was initially presented as a programming technique for eliminating functions with functional arguments~\cite{DBLP:conf/acm/Reynolds72}. The concept and technique generalize to eliminating functions that return functions, for example, the curried version of integer addition, whose type is \texttt{int -> (int -> int)}. The returned functions may contain free variables, and defunctionalization translates free variables into arguments to their corresponding function labels. In general, the label of a function \texttt{f:a -> b} with free variables \texttt{x1:t1, ..., xn:tn} has the form:
\begin{lstlisting}
Fun : t1 -> ... -> tn -> (a, b) lam
\end{lstlisting}
Arguments to the label form an \textit{environment} that stores values of free variables. A label with fully-applied arguments is similar to a \textit{closure}, except that a closure stores the function pointer together with the environment, whereas defunctionalization keeps the function definition separately in \textit{apply}. Example \ref{ex:full dfc} shows the defunctionalization of the curried integer addition.

\begin{exmp}
\label{ex:full dfc}
The source program is the curried integer addition.
\begin{lstlisting}
let plus : int -> (int -> int) = 
    fun x -> (fun y -> x + y)
in
    plus 1 2
\end{lstlisting}
There are two function definitions in the code above. Applying the polymorphic defunctionalization technique explained in Example~\ref{ex:poly dfc}, \texttt{lam} and \texttt{apply} are defined as follows.
\begin{lstlisting}
type ('a, 'b) lam = 
      Plus : (int, (int, int) lam) lam
    | Plus_x : int -> (int, int) lam

let apply : type a b. (a, b) lam -> a -> b =
    fun f arg -> match f with
        | Plus -> Plus_x arg
        | (Plus_x x) -> x + arg
\end{lstlisting}
There is a free variable \texttt{x} of type \texttt{int} in \texttt{(fun y -> x + y)}, so the corresponding constructor \texttt{Plus\_x} takes an argument of type \texttt{int}.

The target program is obtained in a similar way as in previous examples.
\begin{lstlisting}
let plus' : (int, (int, int) lam) lam = Plus
in apply (apply plus' 1) 2
\end{lstlisting}

\end{exmp}

\subsection{Related work in defunctionalization}

Defunctionalization is first presented in its untyped form by Reynolds in the 1970s as a programming technique to translate a higher-order interpreter into a first-order one~\cite{DBLP:conf/acm/Reynolds72}. It is later used as an implementation technique in various applications such as ML compilers~\cite{DBLP:journals/lisp/ChinD96,DBLP:conf/esop/CejtinJW00}, type-safe garbage collectors~\cite{DBLP:conf/popl/WangA01}, and representing higher-kinded polymorphism in OCaml~\cite{DBLP:conf/flops/YallopW14}. The method of using a family of monomorphic \textit{apply} functions to achieve type preservation is the standard workaround for simply-typed defunctionalization in the literature~\cite{DBLP:conf/icfp/BellBH97,DBLP:journals/jfp/TolmachO98,DBLP:conf/esop/CejtinJW00,nielsen2000denotational}. Danvy and Nielsen’s survey contains more examples of defunctionalization in practice~\cite{DBLP:conf/ppdp/DanvyN01}.

% Mention something about Lightweight higher-kinded polymorphism? It's an application of polymorphic DFC.

Formalization of defunctionalization focused on proving type preservation and correctness of the transformation. Bell et al.~proved that the transformation for simply typed programs is type preserving~\cite{DBLP:conf/icfp/BellBH97}.  Nielsen presented a proof for its partial correctness in denotational semantics~\cite{nielsen2000denotational}, and Banerjee et al.~provided a proof for total correctness in operational semantics~\cite{DBLP:conf/tacs/BanerjeeHR01}. Pottier and Gauthier formalized type-preserving polymorphic defunctionalization in System F extended with GADTs~\cite{DBLP:conf/popl/PottierG04}.

\section{The source language (CC)}
\label{sec:2.2}

The source language (which I refer to as CC) I will use to define dependently-typed defunctionalization is a variant of Coquand’s Calculus of Constructions~\cite{DBLP:journals/iandc/CoquandH88}. To be precise, it is a subset of Luo's Extended Calculus of Constructions~\cite{DBLP:phd/ethos/Luo90}, which combines Coquand's original Calculus of Constructions (with only one impredicative universe) with a Martin-Löf style universe hierarchy\footnote{A definition is impredicative if it is self-referencing. The type of an impredicative universe is itself, like \texttt{Type} \texttt{:} \texttt{Type}. In a predicative hierarchy of universes, the type of a universe $U_i$ is always a larger universe $U_{i+1}$. 
}. 

Dependently-typed languages have \textit{universes}, also known as \textit{kinds} or \textit{sorts}, and they are essentially the types of types. CC uses an infinite hierarchy of predicative universes $U_i$, where $i$ ranges over natural numbers, and the type of universe $U_i$ is $U_{i+1}$. Therefore, universes form an infinite hierarchy:
\begin{equation*}
U_0 \mathrel{:} U_1 \mathrel{:} U_2 \mathrel{:} \cdots
\end{equation*}
CC's syntax is presented in Figure~\ref{fig:cc syntax}. Term expressions of CC are: variables (from an infinite set of variable names), universes, dependent function types (or $\Pi$-types), applications, and functions. A CC context is a list of variable-expression pairs.

CC's typing judgements are of the form $\Gamma \vdash e \goodcolon A$, where $\Gamma$ is a context and $e, A$ are terms. A context is \textit{well-formed} if every variable in it is associated with a valid type, that is, the associated expression’s type is a universe. The judgement for the well-formedness of contexts is $\vdash \Gamma$. 

CC's rules for typing (Fig \ref{fig:cc typing}) and well-formedness of contexts (Fig \ref{fig:cc context}) are mutual-inductively defined. The type of a variable $x$ is $A$ if $x \goodcolon A$ is present its well-formed context $\Gamma$ (Var). 
The type of a universe $U_i$ is $U_{i+1}$ (Universe), and the type of $\pitype{x}{A}{B}$ is the highest universe among universes of $A$ and $B$ (Pi). If $e$ has type $B$ in some context $\Gamma$ extended with $x \goodcolon A$, then $\lam{x}{A}{e}$ has the dependent function type $\pitype{x}{A}{B}$ (Lambda). Applications have types $B[e_2\slash x]$, since the output of a function type may depend on the input value (Apply). If an expression $e$ has type $A$ and $A$ is equivalent to $B$ (under context $\Gamma$), then $e$ also has type $B$ (Equiv). 

I write $\Gamma \vdash A \goodcolon U$ to mean that $\Gamma \vdash A \goodcolon U_i$ for some $i$ (which means that $A$ is a type), and $A \rightarrow B$ stands for the $\Pi$-type $\pitype{x}{A}{B}$ where $B$ does not depend on $x$. Base types are omitted for simplicity, but I will use base types like the unit type and the natural numbers freely in examples.

Figure \ref{fig:cc equivalence} defines reduction and equivalence in CC. The only reduction rule is the $\beta$-reduction, $(\lam{x}{A}{e_1}) \ e_2 \ \triangleright \ e_1\sub{e_2}{x}$. I write $e_1 \triangleright^* e_2$ to mean that $e_1$ reduces to $e_2$ in a sequence with zero or more steps. The reduction rule can be thought of the operational semantics of CC.

In CC, two terms are equivalent if they reduce to the same term (Eq-reduce) or are $\eta$-equivalent. The $\eta$-equivalence is defined by two symmetric rules (Eq-Eta1) and (Eq-Eta2). Rule (Eq-Eta1) states that $e_1$ and $e_2$ are equivalent if $e_1$ reduces to a lambda abstraction $\lam{x}{A}{B}$, $e_2$ reduces to some $e_2'$, and $e_2'\ x \equiv e$ in the context extended with $x \goodcolon A$.

\begin{figure}
	\renewcommand{\arraystretch}{1.3}
	\begin{tabular}{l r l l}
		\textit{Universes}   & $U$       & ::= & ${U_i}$ \\
		\textit{Expressions} & ${e}$, ${A}$, ${B}$  & ::= & 
			${x}$ $\ |\ $ ${U}$ $\ |\ $ $\pitype{x}{A}{B}$ $\ |\ $ ${e_1}\ {e_2}$ $\ |\ $ $\lam{x}{A}{e}$\\
		\textit{Context} & ${\Gamma}$ & ::= & ${\cdot}$ $\ |\ $ ${\Gamma}$, ${x} \goodcolon {A}$ \\
	\end{tabular}
	\caption{CC Syntax}
	\label{fig:cc syntax}	
\end{figure}
\begin{figure}
\vspace{0.5cm}
	\begin{equation}
		\tag{Var}
		\frac
			{x \goodcolon A \in \Gamma \qquad \vdash \Gamma}
			{\Gamma \vdash x \goodcolon A}
	\end{equation} \hspace{0.5cm}
	\begin{equation}
		\tag{Universe}
		\frac
			{\vdash \Gamma}
			{\Gamma \vdash U_i \goodcolon U_{i+1}}
	\end{equation} \hspace{0.5cm}
	\begin{equation}
		\tag{Pi}
		\frac
			{\Gamma \vdash A \goodcolon U_i \qquad \Gamma, x \goodcolon A \vdash B \goodcolon U_j
			}
			{\Gamma \vdash \pitype{x}{A}{B} \goodcolon U_{max(i, j)}}
	\end{equation} \hspace{0.5cm}
	\begin{equation}
		\tag{Apply}
		\frac
			{\Gamma \vdash e_1 \mathrel{:} \pitype{x}{A}{B} \qquad \Gamma \vdash e_2 \goodcolon A}
			{\Gamma \vdash e_1 \ e_2 \goodcolon B\sub{e2}{x}}
	\end{equation} \hspace{0.5cm}
	\begin{equation}
		\tag{Lambda}
		\frac
			{ \Gamma, x \goodcolon A \vdash e \goodcolon B}
			{ \Gamma \vdash \lam{x}{A}{e} \mathrel{:} \pitype{x}{A}{B}}
	\end{equation} \hspace{0.5cm}
	\begin{equation}
		\tag{Equiv}
		\frac
			{\Gamma \vdash e \goodcolon A \qquad \Gamma \vdash B \goodcolon U \qquad \Gamma \vdash A \equiv B}
			{\Gamma \vdash e \goodcolon B}
	\end{equation}
	\caption{CC Typing}
	\label{fig:cc typing}
\end{figure}

\begin{figure}
	\begin{equation}
		\tag{WF-Empty}
		\frac{ }{\quad \vdash {\cdot} \quad}
	\end{equation} \hspace{0.5cm}
	\begin{equation}
		\tag{WF-Con}
		\frac
			{\vdash \cdot \qquad \Gamma \vdash A \goodcolon U}
			{\vdash \Gamma , x\goodcolon A}
	\end{equation}
	\caption{Well-formedness of CC context}
    \label{fig:cc context}
\end{figure}

\begin{figure}[H]
	\begin{equation*}
		\tag{Beta}
		(\lam{x}{A}{e_1}) \ e_2 \ \triangleright \ \sub{e_1}{e_2}
	\end{equation*}
	\begin{equation}
		\tag{Eq-reduce}
		\frac
			{\quad e_1 \triangleright^* e \qquad e_ 2\triangleright^* e \quad}
			{\Gamma \vdash e_1 \equiv e_2}
	\end{equation}
	\begin{equation}
		\tag{Eq-Eta1}
		\qquad\qquad\qquad
		\frac
			{e_1 \triangleright^* \lam{x}{A}{e} \qquad
			 e_2 \triangleright^* e_2' \qquad
			 \Gamma, x \goodcolon A \vdash e \equiv e_2'\ x}
			{\Gamma \vdash e_1 \equiv e_2}
	\end{equation}
	\begin{equation}
		\tag{Eq-Eta2}
		\qquad\qquad\qquad
		\frac
			{ e_1 \triangleright^* e_1' \qquad
			 e_2 \triangleright^* \lam{x}{A}{e} \qquad
			 \Gamma, x \goodcolon A \vdash e_1'\ x \equiv e }
			{\Gamma \vdash e_1 \equiv e_2}
	\end{equation}
	\caption{CC reduction and equivalence}
    \label{fig:cc equivalence}
\end{figure}

CC has a useful property: if $\Gamma \vdash e \goodcolon A$, then $\Gamma \vdash A \goodcolon U$. Precisely, in the derivation tree of $\Gamma \vdash e \goodcolon A$, there must be a sub-derivation tree of $\Gamma \vdash A \goodcolon U$.

As standard results from the literature~\cite{DBLP:journals/iandc/CoquandH88,DBLP:phd/ethos/Luo90}, CC is type safe and consistent, and type-checking in CC is decidable. Despite having a minimalistic syntax, CC can express many useful functions, for example, the polymorphic identity function and the dependent composition function.

\begin{exmp}
In System F, the polymorphic identity function has type $\forall a.\ a \rightarrow a$. In CC, we can define a  dependent function that takes a type $A$ and returns the identity function of type $A \rightarrow A$.
	\begin{align*}
		&identity \triangleq \lam{A}{U_0}{\lam{x}{A}{x}} \\
		\cdot \vdash\ &identity\ :\ \pitype{A}{U_0}{A \rightarrow A}
	\end{align*}
\end{exmp}

\begin{exmp}
This example shows that CC can write complicated but meaningful programs like the dependent composition function. The usual composition takes a function $f$ of type $A \rightarrow B$, a function $g$ of type $B \rightarrow C$, an input $x$ of type $A$ (here $A$, $B$, and $C$ are simple types) and returns the result of $g (f\ x)$. The dependent composition generalizes this to the case where $B$ is a family of types indexed by terms of type $A$, and $C$ is a family indexed by terms $x$ of type $A$ and terms of type $B\ x$.
	\begin{align*}
		compose &\triangleq \lam{g}{(\pitype{x}{A}{\pitype{y}{B\ x}{C\ x\ y}})}{\lam{f}{(\pitype{x}{A}{B\ x})}{\lam{x}{A}{g\ x\ (f\ x)}}} \\
		\Gamma &\triangleq \cdot,\ A \goodcolon U_0,\ B \goodcolon A \rightarrow U_0,\ C \goodcolon \pitype{x}{A}{(B\ x) \rightarrow U_0}\\
		\Gamma &\vdash compose \mathrel{:} \pitype{g}{\_}{\pitype{f}{\_}{\pitype{x}{A}{C\ x\ (f\ x)}}} 
	\end{align*}
\end{exmp}


